# DS2-Assignment 1
#rm(list = ls())
# install knitr package
install.packages('knitr')
# load knitr
library(knitr)

```{r setup, include=FALSE} knitr_opts_chunk$set(echo = TRUE, fig.align = 'center', out.width = '80%')```

# Part 1. Artificial data.
## 1.1 Simulation
### 1.1.1 For ARIMA(p,q,d), set
### p = 2
### d = 1
### q = 2 
### phi_1 = .2
### phi_2 = -.15
### theta_1 = .3
### theta_2 = -.1
### noise_sd = 0.03
### and generate a series of sample size n = 1000, using this model.

```{r}
n_size <- 1000
p <- 2
d <- 1
q <- 2
phi_1 <- .2
phi_2 <- -.15
theta_1 <- .3
theta_2 <- -.1
noise_sd <- 0.03

title <- paste0('ARIMA(', p, ', ', d, ', ', q, ')')
                  
set.seed(111)  

sim <- arima.sim(list(order = c (p,d,q), ar = c(phi_1, phi_2), ma = c(theta_1, theta_2)),
                 n = n_size, sd= noise_sd)

plot(sim, xlab = 'Time', ylab = 'Values', main = paste('Simulated', title))
```

#1.1.2 add a linear trend 

```{r}
b0 = -1
b1 = 0.0015
sim_trend <- sim + (b0 + b1 * time(sim))
plot(sim_trend, xlab = 'Time', ylab = 'Values',
     main = paste(title, 'plus linear trend'))
```

#1.1.3 apply an exponential fucntion

```{r}
sim_exp <- exp(sim_trend)
plot(sim_exp, xlab = 'Time', ylab = 'Values', 
     main = paste('exponential function of', arima_title, 'with trend'))
```

## 1.2 Exploratory analysis in the order, inverse to the simulation steps.
### 1.2.1 Divide the generated set into a training set head and a test set tail.

```{r}
ratio <- 0.9
n_size <- length(sim)
train_nsize <- round(n_size * ratio)
# When we take an index slice, it loses ts status.
# We recover it.
train_sim <- ts(sim_exp[1:train_nsize])
cat('Training sample size:', length(train_sim), '\n')
#Same for test
test_sim <- ts(sim_exp[(train_nsize + 1):n_size],
               start = train_nsize + 1 )
cat('Test sample size:', length(test_sim), '\n')
```

### 1.2.2 Logarithm the training set series

```{r}
train_simlog <- log(train_sim)
plot(train_simlog, ylab = 'Values', main = 'Log of training sample', type = 'l')
```

### 1.2.3 Detect a linear trend by regression. Compare the estimated trend parameters to true ones.

```{r}
lm_trend <- lm(train_simlog ~ time(train_simlog))
coef(lm_trend)

trend_par = lm_trend$fitted.values

plot(trend_par, type = 'l', ylab = "Estimated Trend Parameters", main = 'Estimates extracted form training sample')
```

### 1.2.4 Deterend the series.

```{r}
train_detrend <- train_simlog - trend_par
```

### 1.2.5 In the same axis plot the original ARIMA simulation and the current (trended, exponentiated, logarithmed and finally detrended) series. They should have the same shape, but differ by a bit of shift and stretch.

```{r}
plot(sim[1:train_nsize], xlab = 'Time', ylab = 'Values', type = 'l',
     main = 'Simulated vs Log and Detrended training sample',
     ylim = c(min(train_detrend,sim), max(train_detrend, sim)))
lines(train_detrend, col = 'red')
legend('topright', legend = c('Simulated', 'Detrended'), col = c('black', 'red'), lty = 1)
```
### 1.2.6 Fit the ARIMA(2, 1, 2) model (same sizes as the original one)

```{r}
fit_model <- arima(train_detrend, order = c(p, d, q))
```

### 1.2.7 Compare the estimated ARIMA parameteres to true ones. Comment on goodness of fit.
```{r}
fit_model
cat("\nTrue coefficients are phi_1:", phi_1, "phi_2:", phi_2, "theta_1:", theta_1, "theta_2", theta_2, "\n")
```

## 1.3 Forcating the simulated series.
### 1.3.1 Forcats the training part of ARIMA

```{r}
forecast_mdl <- predict(fit_model, n.ahead = 101)
```

### 1.3.2 Add edtimated trend.

```{r}
forecastmdl_trend <- forecast_mdl$pred + (lm_trend$coefficients[1] + lm_trend$coefficients[2] * time(forecast_mdl$pred))
```

### 1.3.3 Exponentiate that trended forecast.
```{r}
forecastmdl_exp <- exp(forecastmdl_trend)
```
 
# 1.4 Evaluation of the results
## 1.4.1 Plot the forecast and real training set in the same axes.

```{r}
forecast_se_exp <- exp(forecast_mdl$se)
upper <- forecastmdl_exp + forecast_se_exp
lower <- forecastmdl_exp - forecast_se_exp

plot(test_sim, type = 'l', 
     main = 'Test time-series vs forecasted with trend, exponent and history',
     ylab = 'Values', xlab = 'Time', xlim = c(890, 1001),
     ylim = c(min(test_sim, forecastmdl_exp, lower),
              max(test_sim, forecastmdl_exp, upper)))
lines(forecastmdl_exp, col = 'red')
lines(ts(sim_exp[890:902],start = 890), col = 'blue')
lines(upper, lty = 2, lwd =2, col = 'springgreen4')
lines(lower, lty = 2, lwd =2, col = 'springgreen4')
abline(v = 902, col = 'darkgreen')
legend('topleft',
       legend = c('real test', 'predicted', 'where forecast starts',
                  'past 10 values', 'confidence interval'),
       col = c('black', 'red', 'darkgreen', 'blue', 'springgreen4'),
       lty = c(rep(1,4),2),cex = .7)
```
### real data
```{r}
plot(sim_exp, type = 'l',
     main = 'Whole simulated time-series vs predicted with trend and exponent',
     ylab = 'Values', xlab = 'Time')
lines(forecastmdl_exp, col = 'red')
legend('topleft', legend = c('simulated', 'predicted'), col = c('black', 'red'),
       lty = 1, cex = 1)
```
### 1.4.2 Plot acf of the training set and its predictors, and ccf between them.

```{r}
acf(test_sim, main = 'ACF of test time-series')
acf(forecastmdl_exp, main = 'ACF of forecast')

ccf(test_sim, forecastmdl_exp, main = 'CCF of test time-series and forecast')
```
### 1.4.3 PLot the residuals and their acf.

```{r}
forecast_res <- test_sim - forecast_exp

plot(forecast_res, type = 'p', ylab = 'Residuals',
     main = 'Residuals plot for forecast')
lines(lowess(forecast_res), col = 'red')
abline(h = 0, lty = 3)

acf(forecast_res, main = 'ACF of residuals')
```

### 1.4.4 Estimate the forecast error.

```{r}
mse <- sum(forecast_res^2) / length(forecast_res)
cat("MSE:", mse, "\n")
cat("Variance of residuals:", var(forecast_res), "\n")
cat("Sum of residuals:", sum(forecast_res), "\n")
```
