# DS2-Assignment 1
#rm(list = ls())
# install knitr package
install.packages('knitr')
# load knitr
library(knitr)

```{r setup, include=FALSE} knitr_opts_chunk$set(echo = TRUE, fig.align = 'center', out.width = '80%')```


#Oil data
##reading the downloaded csv file.

```{r}
#setting the working directory to desktop
setwd("C:\\users\\user\\Desktop\\")

#importing the downloaded csv file
wti <- read.csv("RWTCm.csv")

#creating the date column with desired format 
wti$date <- as.Date(paste(wti$date, '-01', sep = ''),'%B-%Y-%d' )

#we can check to see the nature of imported data by using following commands.
#head(wti)
str(wti)
```
```{r}
# plotting the price
plot(wti$price, type = 'l', xlab = 'time (months)', main = 'Monthly Oil Prices')
legend('topleft', legend = c('Max Price = 133.88 ', 'Min Price = 11.35'))
#plotting the acf for price
acf(wti$price)
```
## 2.2 Exploratory analysis
### 2.2.1 Divide the series into a training set (up to 2019 inclusively) and testing set (all the rest)
```{r}
## there are severla ways to divid the dataset into train and test. th following commented code can also be used.

# we divide the datase in test and train from row 408, where we know that year 2019 ends.
 
#bound = 408
#train <- ( wti[1:bound, ] )             
#test <-  ( wti[(bound+1):nrow(wti), ] )
#train <- ts( wti[1:bound, ] )             
#test <- ts ( wti[(bound+1):nrow(wti), ] )
#head(train)
#tail(train)
#head(test)
#tail(test)

splitSeries <- split(wti,wti$date < as.Date('2020-01-01'))
train <- ts(as.data.frame(splitSeries[2], col.names = '')$price, start = 1986, frequency = 12)

cat ("head of train:\n")
head(train)
cat("\ntail of train:\n")
tail(train)

test <- ts(as.data.frame(splitSeries[1],col.names = '')$price, start = 2020, frequency = 12)
cat ("\nhead of test:\n")
head(test)
cat("\ntail of test:\n")
tail(test)
```
### 2.2.2 Logarithm the series.
# because we can see that there is exponantial behaviour in our data
```{r}
trainLog <- log(train)
```
### 2.2.3 Estimate the linear trend by the least squares procedure.
```{r}
trendFitLog <- lm (trainLog ~ time(trainLog))
# regress logarithm of time series values to time.
coef(trendFitLog)
```
### 2.2.4 Deterend the series.
```{r}
trend <- fitted(trendFitLog)
trainDeterended <- trainLog - trend

plot(trainLog, type = 'l', xlab = 'date', ylab = 'price', main = 'WIT price and extracted trend')
lines(as.vector(time(trainLog)), trend , col = 'Blue')
legend('topleft', legend = c('WTI price', 'trend'), col = c('black','blue'), lty = 1:1, cex = 1.2)

plot(trainDeterended, type ='l', xlab = 'date', ylab = 'price', main = 'Deterended WTI price train set')
```
2.2.5 fit arima(2,1,2) to the training data
```{r}
arimaFit <- arima (trainDeterended, order = c(2,1,2))
arimaFit
```

## 2.3 forcasting in the order, inverse to the exploratory steps
##2.3.1 forcast your arima model for the period of training set.
```{r}
forecast <- predict (arimaFit,n.ahead = length(test))
forecast
```
## 2.3.2 extrapolate your linear trend to this period and add it to your armia forecast.

```{r}
intercept <- trendFitLog$coefficients[1]
slope <- trendFitLog$coefficients[2]
forecastWithTrend <- forecast$pred + intercept +slope *time(forecast$pred)
forecastWithTrend
```
## 2.3.3 exponentiate the result
```{r}
forecastExp <- exp (forecastWithTrend)
forecastExp
```

## 2.4 evaluation of the forecast qulity
## 2.4.1 plot the forecast and test data in the same axis.
```{r}
upper <- forecastExp + exp (forecast$se)
lower <- forecastExp - exp (forecast$se)

plot(test, type = 'l', main = 'Forecasted vs test time-series',
     ylab = 'values', xlab = 'time',ylim = c (10, 70))

lines(forecastExp, col = 'red')
lines(upper, lty = 2, lwd = 2 , col = 'chartreuse')
lines(lower, lty = 2, lwd = 2 , col = 'chartreuse')

abline(v = min (time(test)), col = 'blue')

legend('bottomleft', legend = c('real test', 'predicted', 'where forecast starts', 'confidence interval'), 
       col = c('black', 'red', 'blue', 'chartreuse'), lty = c (rep(1,4),2), cex = 0.7)
```

## 2.3.6 plot the residuals and their acf
##### we know that residuals are real values minus forecast.###
###those residuals sum to zero but these are just inoventions and do not need to sum to zero. so the variance is not the same as mse.###
```{r}
residForecast <- test - forecastExp

plot(residForecast, type = 'p', ylab = 'Residuals', main = 'residuals plot for forecast')
lines(lowess(residForecast), col='red')
abline(h=0 , lty =3 )

acf(residForecast, main = 'ACF of forecast residuals')

```
### 2.3.4 estimate the forecast error
```{r}
mse <- sum (residForecast^2)/ length(residForecast)
cat("Mean Squared Error:", mse, "\n")
sumRes <- sum(residForecast)
cat("Sum of Residuals:", sumRes, "\n")
cat("Variance of Residuals:", var (residForecast))
```